$Z^\prime$ mass limits and the naturalness of supersymmetry

### 2015-03-29 paper/e6ssm-tuning-paper.tex
U(1) --> (1)$

The discovery of a 125 GeV Higgs boson and rising lower bounds on the
masses of superpartners have lead to concerns that supersymmetric
models are now fine tuned.  Large stop masses, required for a $125$
GeV Higgs, feed into the electroweak symmetry breaking conditions 
through renormalisation group equations forcing one to fine tune these 
parameters to obtain the correct electroweak vacuum expectation value.  
Nonetheless this fine tuning depends crucially on our assumptions about 
the supersymmetry breaking scale.  At the same time U(1) extensions
provide the most compelling solution to the $\mu$-problem, which is also 
a naturalness issue, and allow the tree level Higgs mass to be raised substantially above $M_Z$. These very well motivated supersymmetric models 
predict a new $Z^\prime$ boson which could be discovered at the LHC and the
naturalness of the model requires that the $Z^\prime$ boson mass
should not be too far above the TeV scale.  Moreover this fine tuning
appears at the tree level, making it less dependent on assumptions
about the supersymmetry breaking mechanism.  Here we study this fine tuning for
several $U(1)$ supersymmetric extensions of the Standard Model and
compare it to the situation in the MSSM where the most direct tree
level fine tuning can be probed through chargino mass limits.  We show
that future LHC $Z^\prime$ searches are extremely important for
challenging the most natural scenarios in these models.

@@

U(1) extensions of the MSSM provide a very elegant solution to this
$\mu$-problem \cite{Fayet:1977yc, Kim:1983dt, Suematsu:1994qm,
  Cvetic:1995rj, Cvetic:1996mf, Jain:1995cb, Nir:1995bu,
  Cvetic:1997ky} and also raise the Higgs mass with new $F$- and
$D$-terms. Nonetheless, as was recently demonstrated in the context of
the exceptional supersymmetric standard model (E$_6$SSM)
\cite{King:2005jy,King:2005my,Athron:2010zz}, such models can still
suffer from naturalness problems with the mass of the new $Z^\prime$
associated with break down of the new U(1) appearing in the
electroweak symmetry breaking conditions at tree level
\cite{Athron:2013ipa}. Despite this the constrained version of
the E$_6$SSM (cE$_6$SSM) \cite{Athron:2009ue, Athron:2009bs} was still found
to be significantly less tuned than the cMSSM.

@@

At the same time in U(1) extensions, lowering the UV boundary scale
for the RGE evolution also allows even larger $F$-term contributions
to the Higgs mass, so long as one only requires $\lambda$, the
coupling between the Singlet Higgs, $S$ and the up- and down-type
Higgs bosons, $H_u$ and $H_d$, to remain perturbative up to the UV
scale and not the all the way up to the GUT scale.

@@

\section{\label{sec:model}U(1) extensions and the E$_6$SSM}
In this paper we consider U(1) extensions of the MSSM where the gauge
group at low energies is, \be SU(3)_C\times SU(2)_W\times U(1)_Y\times
U(1)^\prime. \ee $U(1)^\prime$ is the new gauge group beyond
that of the SM and MSSM. The minimal superfield content of $U(1)$
extensions which solve the $\mu$-problem should be ordinary left
handed quark, $\hat{Q}_i$ and lepton $\hat{L}_i$
($i=1,2,3$) superfields along with right handed superfields
$\hat{u}^c_i$, $\hat{d}^c_i$, $\hat{e}^c_i$ ($i=1,2,3$) for the
up-type (s)quarks, down-type (s)quarks and charged (s)leptons
respectively and three Higgs superfields, up-type $\hat{H}_u$,
down-type $\hat{H}_d$ and a singlet under the SM gauge group
$\hat{S}$.

@@

Here we will refer to U(1) extensions of the MSSM, which solve the
$\mu$-problem, as the USSM \cite{Cvetic:1995rj, Jain:1995cb,
  Nir:1995bu, Cvetic:1996mf, Cvetic:1997ky}.  The couplings for the
$U(1)^\prime$ gauge group should allow the following renormalisable
superpotential terms required in the USSM, \begin{equation}
 W_{USSM} = y^U_{ij}
\hat{u}^c_i  \hat{H}_u \cdot \hat{Q}_j + y^D_{ij} \hat{d}^c_i  \hat{Q}_j \cdot \hat{H}_d
 + y^E_{ij} \hat{e}^c_i \hat{L}_j \cdot \hat{H}_d  + \lambda \hat{S} \hat{H}_d \cdot \hat{H}_u, 
\end{equation} with $i,j \in \{1,2,3\}$. 
For the $SU(2)$ dot product we follow the convention $\hat{A}\cdot \hat{B}\equiv \epsilon_{\alpha\beta}\hat{A}^\alpha\hat{B}^\beta=\hat{A}^2\hat{B}^1-\hat{A}^1\hat{B}^2$.

@@

The exact level of tuning from the $Z^\prime$ depends on the charges
of the extra $U(1)$ gauge symmetry it is associated with.  In
Fig.~\ref{Fig:othere6modelsvsmssm} we look at the fine tuning for
other U(1) extensions for the same $Z^\prime$ masses as we did for the
E$_6$SSM.  To simplify the analysis we fix $\tan \beta = 10$, but scan
over the remaining parameters as in Table \ref{tab:scanranges} and fix
the rest to the same values we did in the scan carried out for
Fig.~\ref{Fig:e6ssmvsmssm}. In order to more clearly identify the lower bound
on the obtainable tuning in each model, the parameter values for points in
these main grid scans with a low fine tuning were then used as the starting points for
smaller scans about those values. In these smaller scans the parameters were more finely
varied to populate the low fine tuning regions.

@@

We have investigated this question here in the context of the MSSM and
U(1) extensions.  Since the RG evolution links the soft masses
together and causes these problems from stop and gluino masses the
most conservative approach to placing naturalness limits is to choose
a low cutoff. We find that in the MSSM the most direct way to
constrain naturalness of the model, without making assumptions about
the SUSY breaking scale is through limits on the chargino
masses. Current LHC limits on charginos are not model independent and
thereby leave many gaps where one can have light charginos.

>>>>

The discovery of a 125 GeV Higgs boson and rising lower bounds on the
masses of superpartners have lead to concerns that supersymmetric
models are now fine tuned.  Large stop masses, required for a $125$
GeV Higgs, feed into the electroweak symmetry breaking conditions 
through renormalisation group equations forcing one to fine tune these 
parameters to obtain the correct electroweak vacuum expectation value.  
Nonetheless this fine tuning depends crucially on our assumptions about 
the supersymmetry breaking scale.  At the same time $U(1)$ extensions
provide the most compelling solution to the $\mu$-problem, which is also 
a naturalness issue, and allow the tree level Higgs mass to be raised substantially above $M_Z$. These very well motivated supersymmetric models 
predict a new $Z^\prime$ boson which could be discovered at the LHC and the
naturalness of the model requires that the $Z^\prime$ boson mass
should not be too far above the TeV scale.  Moreover this fine tuning
appears at the tree level, making it less dependent on assumptions
about the supersymmetry breaking mechanism.  Here we study this fine tuning for
several $U(1)$ supersymmetric extensions of the Standard Model and
compare it to the situation in the MSSM where the most direct tree
level fine tuning can be probed through chargino mass limits.  We show
that future LHC $Z^\prime$ searches are extremely important for
challenging the most natural scenarios in these models.

@@

$U(1)$ extensions of the MSSM provide a very elegant solution to this
$\mu$-problem \cite{Fayet:1977yc, Kim:1983dt, Suematsu:1994qm,
  Cvetic:1995rj, Cvetic:1996mf, Jain:1995cb, Nir:1995bu,
  Cvetic:1997ky} and also raise the Higgs mass with new $F$- and
$D$-terms. Nonetheless, as was recently demonstrated in the context of
the exceptional supersymmetric standard model (E$_6$SSM)
\cite{King:2005jy,King:2005my,Athron:2010zz}, such models can still
suffer from naturalness problems with the mass of the new $Z^\prime$
associated with break down of the new $U(1)$ appearing in the
electroweak symmetry breaking conditions at tree level
\cite{Athron:2013ipa}. Despite this the constrained version of
the E$_6$SSM (cE$_6$SSM) \cite{Athron:2009ue, Athron:2009bs} was still found
to be significantly less tuned than the cMSSM.

@@

At the same time in $U(1)$ extensions, lowering the UV boundary scale
for the RGE evolution also allows even larger $F$-term contributions
to the Higgs mass, so long as one only requires $\lambda$, the
coupling between the Singlet Higgs, $S$ and the up- and down-type
Higgs bosons, $H_u$ and $H_d$, to remain perturbative up to the UV
scale and not the all the way up to the GUT scale.

@@

\section{\label{sec:model}$U(1)$ extensions and the E$_6$SSM}
In this paper we consider $U(1)$ extensions of the MSSM where the gauge
group at low energies is, \be SU(3)_C\times SU(2)_W\times U(1)_Y\times
U(1)^\prime. \ee $U(1)^\prime$ is the new gauge group beyond
that of the SM and MSSM. The minimal superfield content of $U(1)$
extensions which solve the $\mu$-problem should be ordinary left
handed quark, $\hat{Q}_i$ and lepton $\hat{L}_i$
($i=1,2,3$) superfields along with right handed superfields
$\hat{u}^c_i$, $\hat{d}^c_i$, $\hat{e}^c_i$ ($i=1,2,3$) for the
up-type (s)quarks, down-type (s)quarks and charged (s)leptons
respectively and three Higgs superfields, up-type $\hat{H}_u$,
down-type $\hat{H}_d$ and a singlet under the SM gauge group
$\hat{S}$.

@@

Here we will refer to $U(1)$ extensions of the MSSM, which solve the
$\mu$-problem, as the USSM \cite{Cvetic:1995rj, Jain:1995cb,
  Nir:1995bu, Cvetic:1996mf, Cvetic:1997ky}.  The couplings for the
$U(1)^\prime$ gauge group should allow the following renormalisable
superpotential terms required in the USSM, \begin{equation}
 W_{USSM} = y^U_{ij}
\hat{u}^c_i  \hat{H}_u \cdot \hat{Q}_j + y^D_{ij} \hat{d}^c_i  \hat{Q}_j \cdot \hat{H}_d
 + y^E_{ij} \hat{e}^c_i \hat{L}_j \cdot \hat{H}_d  + \lambda \hat{S} \hat{H}_d \cdot \hat{H}_u, 
\end{equation} with $i,j \in \{1,2,3\}$. 
For the $SU(2)$ dot product we follow the convention $\hat{A}\cdot \hat{B}\equiv \epsilon_{\alpha\beta}\hat{A}^\alpha\hat{B}^\beta=\hat{A}^2\hat{B}^1-\hat{A}^1\hat{B}^2$.

@@

The exact level of tuning from the $Z^\prime$ depends on the charges
of the extra $U(1)$ gauge symmetry it is associated with.  In
Fig.~\ref{Fig:othere6modelsvsmssm} we look at the fine tuning for
other $U(1)$ extensions for the same $Z^\prime$ masses as we did for the
E$_6$SSM.  To simplify the analysis we fix $\tan \beta = 10$, but scan
over the remaining parameters as in Table \ref{tab:scanranges} and fix
the rest to the same values we did in the scan carried out for
Fig.~\ref{Fig:e6ssmvsmssm}. In order to more clearly identify the lower bound
on the obtainable tuning in each model, the parameter values for points in
these main grid scans with a low fine tuning were then used as the starting points for
smaller scans about those values. In these smaller scans the parameters were more finely
varied to populate the low fine tuning regions.

@@

We have investigated this question here in the context of the MSSM and
$U(1)$ extensions.  Since the RG evolution links the soft masses
together and causes these problems from stop and gluino masses the
most conservative approach to placing naturalness limits is to choose
a low cutoff. We find that in the MSSM the most direct way to
constrain naturalness of the model, without making assumptions about
the SUSY breaking scale is through limits on the chargino
masses. Current LHC limits on charginos are not model independent and
thereby leave many gaps where one can have light charginos.
